#htseq-count input
```{r}
directory <- "/Users/samuellu/Desktop/GDC/all"
```

```{r}
sampleFiles <- grep("group",list.files(directory),value=TRUE)
sampleCondition <- sub("(.*group).*","\\1",sampleFiles)
sampleTable <- data.frame(sampleName = sampleFiles,
                          fileName = sampleFiles,
                          condition = sampleCondition)
sampleTable$condition <- factor(sampleTable$condition)
```

It may take a while.
```{r}
library("DESeq2")
dds <- DESeqDataSetFromHTSeqCount(sampleTable = sampleTable,
                                       directory = directory,
                                       design= ~ condition)
dds
```
#Pre-filtering
remove rows in which there are reads less than 10.
```{r}
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]
```
#Note on factor levels
tell results which comparison to make.
```{r}
dds$condition <- factor(dds$condition, levels = c("younggroup","oldgroup"))
#dds$condition
```
#Differential expression analysis
It may take a while.
```{r}
dds <- DESeq(dds)
res <- results(dds)
res
```
specify the contrast I want to build a results table.
```{r}
res <- results(dds, contrast=c("condition","younggroup","oldgroup"))
res
```
#Log fold change shrinkage for visualization and ranking
```{r}
resultsNames(dds)
```

```{r}
library(apeglm)
resLFC <- lfcShrink(dds, coef="condition_oldgroup_vs_younggroup", type="apeglm")
resLFC
```
#Speed-up and parallelization thoughts
```{r}
library("BiocParallel")
register(MulticoreParam(4))
```
#p-values and adjusted p-values
```{r}
resOrdered <- res[order(res$pvalue),]
```

summarize some basic tallies using the summary function.
```{r}
summary(res)
```
to check how many adjusted p-values were less than 0.1.
```{r}
sum(res$padj < 0.1, na.rm=TRUE)
```
set the adjusted p value cutoff to 0.05.
```{r}
res05 <- results(dds, alpha=0.05)
summary(res05)
```

```{r}
sum(res05$padj < 0.05, na.rm=TRUE)
```
#Independent hypothesis weighting
implement the method of Independent Hypothesis Weighting.
```{r}
library("IHW")
resIHW <- results(dds, filterFun=ihw)
summary(resIHW)
sum(resIHW$padj < 0.1, na.rm=TRUE)
metadata(resIHW)$ihwResult
```
#Exploring and exporting results
##MA-plot
```{r}
plotMA(res, ylim=c(-2,2))
```
remove the noise associated with log2 fold changes from low count genes without requiring arbitrary filtering thresholds.
```{r}
plotMA(resLFC, ylim=c(-2,2))
```

#Alternative shrinkage estimators
```{r}
resultsNames(dds)
```
# because we are interested in younggroup vs oldgroup, we set 'coef=2'
It may take a while.
```{r}
resNorm  <- lfcShrink(dds, coef=2, type="normal")
resAsh <- lfcShrink(dds, coef=2, type="ashr")
```

```{r}
par(mfrow=c(1,3), mar=c(4,4,2,1))
xlim <- c(1,1e5); ylim <- c(-3,3)
plotMA(resLFC, xlim=xlim, ylim=ylim, main="apeglm")
plotMA(resNorm, xlim=xlim, ylim=ylim, main="normal")
plotMA(resAsh, xlim=xlim, ylim=ylim, main="ashr")
```

#Plot counts
examine the counts of reads for a single gene across the groups.
```{r}
plotCounts(dds, gene=which.min(res$padj), intgroup="condition")
```
an argument returnData specifies that the function should only return a data.frame for plotting with ggplot.
```{r}
d <- plotCounts(dds, gene=which.min(res$padj), intgroup="condition", 
                returnData=TRUE)
library("ggplot2")
ggplot(d, aes(x=condition, y=count)) + 
  geom_point(position=position_jitter(w=0.1,h=0)) + 
  scale_y_log10(breaks=c(25,100,400))
```
#More information on results columns
```{r}
mcols(res)$description
```
#Exporting results to CSV files
A plain-text file of the results can be exported using the base R functions write.csv.
```{r}
write.csv(as.data.frame(resOrdered), 
          file="condition_age_results.csv")
```
Exporting only the results which pass an adjusted p value threshold can be accomplished with the subset function, followed by the write.csv function.
```{r}
resSig_d <- subset(resOrdered, padj < 0.1)
resSig_d[which(resSig_d$log2FoldChange > 0), "gen_reg"] <- "up"
resSig_d[which(resSig_d$log2FoldChange < 0), "gen_reg"] <- "down"
resSig_d
```
#remove ensembl_id "."
```{r}
ens_id<- substr(row.names(resSig_d),1 ,15)
rownames(dds) <- ens_id

```

```{r}
dds
```

```{r}
write.csv(as.data.frame(resSig_d), 
          file="condition_age_results_resSig_d.csv")
```

#Data transformations and visualization
##Count data transformations
###Extracting transformed values
```{r}
vsd <- vst(dds, blind=FALSE)
head(assay(vsd), 3)
```
# this gives log2(n + 1)
```{r}
ntd <- normTransform(dds)
library("vsn")
meanSdPlot(assay(ntd))
```

```{r}
meanSdPlot(assay(vsd))
```

```{r}
#meanSdPlot(assay(rld))
```
#Data quality assessment by sample clustering and visualization
##Heatmap of the count matrix!
```{r}
res
```

Below we show how to produce such a heatmap for various transformations of the data.
I remove the "annotation_col=df" and "show_rownames=FALSE", then add "cluster_rows=FALSE" and "cluster_cols=FALSE" to make the plot clearer.
```{r}
library("pheatmap")
select <- order(rowMeans(counts(dds,normalized=TRUE)),
                decreasing=TRUE)[1:20]
df <- as.data.frame(colData(dds)[,c("condition")])

pheatmap(assay(ntd)[select,], cluster_rows=FALSE, cluster_cols=FALSE, show_rownames = F, show_colnames = F) #, annotation_col=df)
```

```{r}
pheatmap(assay(vsd)[select,], cluster_rows=FALSE, cluster_cols=FALSE, show_rownames=FALSE, show_colnames = F)#,cluster_cols=FALSE, annotation_col=df)
```

##Heatmap of the sample-to-sample distances
get sample-to-sample distances.
just wait for a while.
```{r}
sampleDists <- dist(t(assay(vsd)))
```

```{r}
library("RColorBrewer")
sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- paste(vsd$condition, vsd$type, sep="-")
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
pheatmap(sampleDistMatrix,
         clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists,
         col=colors, show_rownames=FALSE)
```

##Principal component plot of the samples
shows the samples in the 2D plane spanned by their first two principal components. 
```{r}
plotPCA(vsd, intgroup=c("condition"))
```

customize the PCA plot using the ggplot function.
```{r}
pcaData <- plotPCA(vsd, intgroup=c("condition"), returnData=TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))
ggplot(pcaData, aes(PC1, PC2, color=condition, shape=condition)) +
  geom_point(size=2) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed()

```
